---
title: 'Final Exam Markdown: Group 1'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
```

The fall 2020 IST722 final exam has two parts: a statistical analysis you develop in class
and a report that you write on your own time. The first part, which you are working on
right now, involves using this Markdown file to produce all of the diagnostics, graphs, and
statistical output you will need for your report.

The **Knit** button at the top of the code window will generate an html file that
includes both content as well as the output of any embedded R code chunks within this 
document. When you open this html file in a browser, you will be able to save it as
a pdf file for submission to Blackboard. The file you submit to Blackboard must contain
all of the output that you will use in your report. Any questions you answer with results
or graphics not appearing in your PDF will not receive credit.

This initial code block opens the file and produces basic diagnostics. Make sure you have
set the working directory and the file name correctly.

```{r open}
# First, set the working directory to wherever you have stored your data file.
# setwd("~/Downloads")
# Already stored the markdown file in the same directory. So, I won't be setting the 
# working directory

# In this block, modify the file name of the Rdata file to match the name of
# the file you downloaded from Blackboard. Change the 1 to whatever number you have
load("datasets15.RData")

# If you can't get that to work, comment out the previous line and try file.choose() to select 
# your data file using a file selection dialog:
# 
# load(file.choose())

# Now make sure that the two datasets you will be using are available.
str(usVaccines)
dim(usVaccines)
str(districts)
dim(districts)

```




## Plotting the Time Series Data

A basic time series plot can be valuable for examining trends,
periodicity, and other aspects of a time series. You don't necessarily need to include
this plot in your report, particularly because you only care about vaccination levels
in the most recent few years.

```{r tsplot, echo=TRUE}

plot(usVaccines)

# Key to abbreviations (from WHO)
# DTP - Diptheria, Tetanus, Pertussis
# HepB_BD - Hepatitis B, Birth Dose
# Pol3 - Inactivated polio vaccine
# Hib3 - Haemophilus influenza type B vaccine
# MCV - Measles-containing-vaccine first-dose

```


## Time Series Autocorrelation

One of the most basic diagnostics for time series is an autocorrelation plot. One of
the exam questions asks you to interpret these plots. You need not modify this block 
of code, as it produces all five of the plots you need.


```{r TSacf, echo=TRUE}

# This runs an ACF on each of the five time series 
for (i in 1:5) {
  acf(usVaccines[,i], main=attr(usVaccines,"dimnames")[[2]][i])
}

for (i in 1:5) {
  plot(decompose(ts(usVaccines[, i], frequency = 2)))
}

```

## Time Series Changepoint Analysis of Means

This code calculates a changepoint analysis of means that is designed to
find the last mean changepoint within each of the five series. As long as you have
the changepoint package installed, there should be no need to modify this code.

```{r TScpt, echo=TRUE}
library(changepoint)

# This finds the final mean changepoint for each time series
print("Last mean changepoint for each time series:")
for (i in 1:5) {
  print(attr(usVaccines,"dimnames")[[2]][i])
  print(changepoint::cpt.mean(rev(usVaccines[,i]), method="AMOC")@cpts[1])
}

```


## Time Series Subsetting

You will be calibrating your district-by-district vaccine results by comparing to
the U.S. mean level of vaccination during recent years. Subset the time series data 
to only include the most recent years during which all vaccine mean levels were 
stable. Use the changepoint output from the previous step to make this judgment call. 
Set the value of startPoint to the most appropriate starting point within the 38 year
period of the time series data. Then calculate the mean level for each vaccination 
during that time period. 

Optionally add a bar plot that puts all of the mean vaccination levels into the same graphic.

```{r TSsubset, echo=TRUE}

startPoint <- 14 # Change the starting point for calculating a mean of recent observations only.
# Choose the most reasonable value by looking at the results of the previous code block.

# Now calculate the mean vaccination level during recent years for
# each type of vaccine. 
mean_vacc <- apply(usVaccines[startPoint:38,], MARGIN=2, FUN=mean)

# Optionally add a barplot() to display these means.
barplot(mean_vacc)

```

That's the end of the analysis of the usVaccines data. Next, you will conduct diagnostics
on your main dataset, the district by district results for a random sample of n=700 California
school districts. Refer to the exam specification for information on the meaning of
each variable in your data set.

## Descriptive Statistics for Later Use

Clients expect basic descriptive statistics on the major variables of importance. Descriptives
may also be important in guiding your analyses. The code
below produces a summary of the data set, but you may also want to compute some 
other statistics that do not appear in the output of summary().

```{r descriptives, echo=TRUE}

summary(districts)
# Add any additional descriptive statistics that you want to view now or later.

# This produces a complete correlation matrix, rounded to two decimal digits
round(cor(districts[,-1]),2)
round(cor(districts[,2:5]),2)
colMeans(districts[2:5])
aggregate(districts[,2:5],by=list(districts$DistrictName), mean)
mean(districts$PctUpToDate)
```


## Plotting Histograms as Diagnostics

Most reports that you create for clients will NOT benefit from including histograms
because histograms are very low level and rarely communicate much of interest.
Nonetheless, you should look at them yourself to make sure you understand your data.
Add any additional histograms that you wish to view.

```{r histograms, echo=TRUE}

hist(districts$TotalSchools)
# Add any additional histograms that you want to view now or later.
hist(districts$PctUpToDate)
hist(districts$WithoutDTP)
hist(districts$WithoutPolio)
hist(districts$WithoutMMR)
hist(districts$WithoutHepB)
hist(districts$PctBeliefExempt) # 2
hist(districts$PctChildPoverty)
hist(districts$PctFamilyPoverty)
hist(districts$Enrolled) # 3

```


## Transforming Variables Prior to Inferential Analysis

Based on the histograms you created above, you may want to make some transformations to
your numeric variables to reshape their respective distributions. Remember to
add new variables to your dataset, rather than overwriting existing variables.

```{r transformations, echo=TRUE}

# Here's an example: The total number of schools in each district is generally small
# but there are a few large districts. Try a log() transformation to address this.
districts$logTotSchools <- log(districts$TotalSchools)
hist(districts$logTotSchools)

# Add any additional transformations that you want to compute and store on your data set.
# Generally speaking, you should inspect the results of each transformation with a histogram
# or some descriptive statistics.

districts$logPctBeliefExempt <- log(districts$PctBeliefExempt)
hist(districts$logPctBeliefExempt)

districts$logEnrolled <- log(districts$Enrolled)
hist(districts$logEnrolled)

```

## Inferential Analyses

One very important aspect of doing this final exam is that you need to anticipate what kinds of
analytical results will go into your report. Now is a good time to reexamine the
exam specification to make sure that you know what analyses you will need in order to 
complete your report. For example, the following code produces results for a
t-test which you may or may not need. Subsequent code blocks create linear regression
and logistic regression results. Add any additional t-tests that you think you may need 
in this block of code.

```{r ttests, echo=TRUE}
library(BayesFactor)

# These t-tests compare the mean number of schools per district for those
# districts that completed reporting versus those that did not.
t.test(TotalSchools ~ DistrictComplete, data=districts)
ttestBF(formula=TotalSchools ~ DistrictComplete, data=districts, posterior=FALSE)
summary(ttestBF(formula=TotalSchools ~ DistrictComplete, data=districts, posterior=TRUE, iterations=10000))
# Add any additional t-tests that you wish to compute.


```

```{r}
summary(lm(PctUpToDate ~ WithoutDTP+WithoutMMR+WithoutHepB+WithoutPolio+PctBeliefExempt, data=districts))
```

The code below produces a linear regression - a one predictor model that predicts
the percentage of students who are up to date based on the total student enrollment.
This code produces a frequentist result, a Bayes Factor, and a Bayesian estimation
result. Add any additional regressions you would like to compute by copying the code
and adding new/different predictors.

```{r regressions, echo=TRUE}
library(BayesFactor)


# These regressions predict the percentage of students who are up to date on vaccines
# using the number of students enrolled in the district.
summary(lm(PctBeliefExempt ~ logEnrolled+PctChildPoverty+PctFamilyPoverty+logTotSchools, data=districts))
summary(lmBF(PctBeliefExempt ~ logEnrolled+PctChildPoverty+PctFamilyPoverty+logTotSchools, data=districts, posterior=FALSE))
summary(lmBF(PctBeliefExempt ~ logEnrolled+PctChildPoverty+PctFamilyPoverty+logTotSchools, data=districts, posterior=TRUE, iterations=10000))
# Add any additional regression analyses that you wish to compute
```


```{r regressions1, echo=TRUE}
summary(lm(PctUpToDate ~ logEnrolled+PctChildPoverty+PctFamilyPoverty+logTotSchools, data=districts))
summary(lmBF(PctUpToDate ~ logEnrolled+PctChildPoverty+PctFamilyPoverty+logTotSchools, data= districts, posterior=FALSE))
summary(lmBF(PctUpToDate ~ logEnrolled+PctChildPoverty+PctFamilyPoverty+logTotSchools, data=districts, posterior=TRUE, iterations=10000))
```

```{r}
summary(lm(PctUpToDate ~ PctChildPoverty*Enrolled, data = districts))
summary(lm(PctUpToDate ~ PctChildPoverty+Enrolled, data = districts))
summary(lmBF(PctUpToDate ~ PctChildPoverty*Enrolled, data=districts, posterior=FALSE))
summary(lmBF(PctUpToDate ~ PctChildPoverty+Enrolled, data=districts, posterior=FALSE))
```


```{r regressions2, echo=TRUE}
summary(lmBF(PctUpToDate ~ logEnrolled+PctChildPoverty+PctFamilyPoverty+logTotSchools, data=districts, posterior=FALSE))
summary(lmBF(PctUpToDate ~ logEnrolled+PctChildPoverty+PctFamilyPoverty+logTotSchools, data=districts, posterior=TRUE, iterations=10000))
```

Finally, the code below produces a logistic regression - a one predictor model that predicts
the percentage of students who are up to date based on the total student enrollment. The
code produces a frequentist result as well as a Bayesian estimation result.
Add any additional regressions you would like to compute by copying the code
and adding new/different predictors.

```{r logistic, echo=TRUE}

library(MCMCpack)


# These logistic regressions predict whether a district's reporting is complete
# based on the percentage of students who did not get the DTP vaccine. 
glmOut <- glm(DistrictComplete ~ PctFamilyPoverty+logEnrolled+ logTotSchools , family = binomial(), data=districts)
summary(glmOut)
exp(coef(glmOut))
exp(confint(glmOut))
glmBayesOut <- MCMCpack::MCMClogit(DistrictComplete ~ PctFamilyPoverty+logEnrolled+ logTotSchools , data=districts)
summary(glmBayesOut)
summary(exp(glmBayesOut))
# Add any additional logistic regression analyses that you wish to compute

glmOut <- glm(DistrictComplete ~ logEnrolled+ logTotSchools , family = binomial(), data=districts)
summary(glmOut)
exp(coef(glmOut))
exp(confint(glmOut))
glmBayesOut <- MCMCpack::MCMClogit(DistrictComplete ~ logEnrolled+ logTotSchools, data=districts)
summary(glmBayesOut)
summary(exp(glmBayesOut))
```


## Knit, PDF, and Submit to Blackboard

Click the **Knit** button to create and inspect your html document. If the knitting process 
is successful an html file will be submitted to your current working directory. 
The file will also pop up in a viewer. There is a button on the viewer that says "Open in
Browser." Click this and then save a PDF version of your page from your browser. Submit
the PDF version to Blackboard (Week 13 Content).